On Optimality of Finite Element Integration

### 2015-11-06 TOMS_Paper/assembly.tex
With Francis minor fixes

As shown later, multilinear loop nests are important because they have a structure that we can take advantage of to synthesize optimal code. We will see that they arise naturally when translating bilinear or linear forms into code.

Two other classes of loops need be defined.

@@

\begin{Def}[Optimality of a loop nest]
\label{def:mln-optimality}
Let $\Lambda$ be a generic loop nest, and let $\Gamma$ be a generic transformation function $\Gamma : \Lambda \rightarrow \Lambda'$ such that $\Lambda'$ is semantically equivalent to $\Lambda$ (possibly, $\Lambda' = \Lambda$). We say that $\Lambda' = \Gamma (\Lambda)$ is an optimal synthesis of $\Lambda$ if the number of operations (sums, products) that it performs to evaluate the result is minimal.
\end{Def}

@@

To synthesize optimal implementations as in Definition~\ref{def:mln-optimality} we need:
\begin{itemize}
\item a characterization of the space of legal transformations impacting the operation count
\item a cost model to select the optimal point in the transformation space
\end{itemize}
In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we focus on the construction of such transformation space. We consider multilinear loop nests of arity $n=2$ (i.e., bilinear forms), although a similar (simpler) reasoning straightforwardly applies to the case $n < 2$.

@@

\subsubsection{Reduction and free order loops}
\label{sec:se-rln}
Integration- and geometry-dependent expressions should logically be placed, respectively, within the reduction and free order loops. The factorization of $\sigma_{ei}$, therefore, also plays a key role for optimal sharing elimination. This is particularly important in complex forms, characterized by several coefficients, derivatives of coefficients, and extensive use of differential operators. We distinguish two cases:
\begin{itemize}
\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the normal form reduces to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus f_i$. 
\item There is no common symbol among all terms. There still can be two terms $w_1$ and $w_2$ such that $f_i \in \sigma_{ei}^{w1}, \sigma_{ei}^{w2}$, but factorizing $f_i$ would only increase the operation count since sharing elimination opportunities across test and trial functions would be lost. On the other hand, sharing elimination must occur \textit{within} each term $\sigma_{ei}^{w}$: factorization of (derivatives of) coefficients promotes reduction-loop invariant code motion. 
\end{itemize}

\subsubsection{Observations and underlying assumptions}
\label{sec:se-assumptions}
We observe that there almost never are two ore more conflicting factorizations among which to choose for minimizing the operation count. The idea is that, by choosing a factorization that places a sub-expression within the loop that it logically belongs to, a reduction in operation count is generally obtained. There only are two pathological circumstances in which such a reduction may not be obtained: 
\begin{itemize}
\item As explained in Section~\ref{sec:se-opln}, having both test and trial functions in the same function space exposes sub-expressions over different loops, which introduces a choice between two alternative transformations.
\item The trip count of a loop is extremely small (typically it needs be 1 for the following to happen). This decreases or nullifies the impact of code motion -- exposed through factorization, as described in Sections~\ref{sec:se-opln} and~\ref{sec:se-rln}. For instance, the reader can verify that expanding the expression $(a f_i + b g_i) (c f_i + d g_i)$ within a loop $L_i$\footnote{The example was excerpted from the multiplication of the Jacobian matrix for the change of coordinates with the derivative of a coefficient} and factorizing $L_i$-dependent symbols leads to optimality if and only if the size of $L_i$ is strictly greater than 1.
\end{itemize}

@@

\begin{Def}[Optimality of a loop nest with bounded working set]
\label{def:optimality}
Let $\Lambda$ be a generic loop nest, and let $\Gamma$ be a generic transformation function $\Gamma : \Lambda \rightarrow \Lambda'$ such that $\Lambda'$ is semantically equivalent to $\Lambda$ (possibly, $\Lambda' = \Lambda$) and Constraints~\ref{const:Le} and~\ref{const:TH} are satisfied. We say that $\Lambda' = \Gamma (\Lambda)$ is an optimal synthesis of $\Lambda$ if the number of operations (sums, products) that it performs to evaluate the result is minimal.
\end{Def}

>>>>

We will show that multilinear loop nests, which arise naturally when translating bilinear or linear forms into code, are important because they have a structure that we can take advantage of to synthesize optimal code.

We define two other classes of loops.

@@

\begin{Def}[Optimality of a loop nest]
\label{def:mln-optimality}
Let $\Lambda$ be a generic loop nest, and let $\Gamma$ be a generic transformation function $\Gamma : \Lambda \rightarrow \Lambda'$ such that $\Lambda'$ is semantically equivalent to $\Lambda$ (possibly, $\Lambda' = \Lambda$). We say that $\Lambda' = \Gamma (\Lambda)$ is an optimal synthesis of $\Lambda$ if the number of operations (additions, products) that it performs to evaluate the result is minimal.
\end{Def}

@@

To synthesize optimal implementations as in Definition~\ref{def:mln-optimality} we need:
\begin{itemize}
\item a characterization of the space of legal transformations impacting the operation count
\item a cost model to select the optimal point in the transformation space
\end{itemize}
In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we focus on the construction of such a transformation space. We consider multilinear loop nests of arity $n=2$ (i.e., bilinear forms), although a similar (simpler) reasoning straightforwardly applies to the case $n < 2$.

@@

\subsubsection{Reduction and free order loops}
\label{sec:se-rln}
Integration- and geometry-dependent expressions should logically be placed, respectively, within the reduction and free order loops. The factorization of $\sigma_{ei}$, therefore, also plays a key role for optimal sharing elimination. This is particularly important in complex forms, characterized by several coefficients, derivatives of coefficients, and extensive use of differential operators. We distinguish two cases:
\begin{itemize}
\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the normal form reduces to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus \lbrace f_i \rbrace$. 
\item There is no common symbol among all terms. There still can be two terms $w_1$ and $w_2$ such that $f_i \in \sigma_{ei}^{w1}, \sigma_{ei}^{w2}$, but factorizing $f_i$ would only increase the operation count since sharing elimination opportunities across test and trial functions would be lost. On the other hand, sharing elimination must occur \textit{within} each term $\sigma_{ei}^{w}$: factorization of (derivatives of) coefficients promotes reduction-loop invariant code motion. 
\end{itemize}

\subsubsection{Observations and underlying assumptions}
\label{sec:se-assumptions}
We observe that there are almost never two ore more conflicting factorizations among which to choose for minimizing the operation count. The idea is that, by choosing a factorization that places a sub-expression within the loop that it logically belongs to, a reduction in operation count is generally obtained. There only are two pathological circumstances in which such a reduction may not be obtained: 
\begin{itemize}
\item As explained in Section~\ref{sec:se-opln}, having both test and trial functions in the same function space exposes sub-expressions over different loops, which introduces a choice between two alternative transformations.
\item The trip count of a loop is extremely small (typically it needs be 1 for the following to happen). This decreases or nullifies the impact of code motion -- exposed through factorization, as described in Sections~\ref{sec:se-opln} and~\ref{sec:se-rln}. For instance, the reader can verify that expanding the expression $(a f_i + b g_i) (c f_i + d g_i)$ within a loop $L_i$\footnote{The example was excerpted from the multiplication of the Jacobian matrix for the change of coordinates with the derivative of a coefficient} and factorizing $L_i$-dependent symbols leads to optimality if and only if the size of $L_i$ is strictly greater than 1.
\end{itemize}

@@

\begin{Def}[Optimality of a loop nest with bounded working set]
\label{def:optimality}
Let $\Lambda$ be a generic loop nest, and let $\Gamma$ be a generic transformation function $\Gamma : \Lambda \rightarrow \Lambda'$ such that $\Lambda'$ is semantically equivalent to $\Lambda$ (possibly, $\Lambda' = \Lambda$) and Constraints~\ref{const:Le} and~\ref{const:TH} are satisfied. We say that $\Lambda' = \Gamma (\Lambda)$ is an optimal synthesis of $\Lambda$ if the number of operations (additions, products) that it performs to evaluate the result is minimal.
\end{Def}
